Modeling the competing effects of the immune system and EMT on epithelial cancers

### 2019-09-17 Paper/Paper.tex
Added LIHC and PAAD to Paper and CleanPaper

\tcr{In addition, the immune system plays important regulatory roles in pre-cancerous tissue, where inflammation can increase the probability of tumorigenesis (and/or decrease the time to cancer), \sout{particularly} \tcb{including} in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.}

@@

\tcb{There is evidence directly linking Treg-secreted TGF-$\beta$ with EMT in hepatocellular carcinoma\cite{shi2019cd4+}.} Thus, even by considering only a single signaling factor, we find these three components (the tumor, the immune system, and EMT) to be linked. It therefore strikes us as a priority to develop models to understand how interactions between each of these three components affect cancer incidence and progression.

@@

We seek to describe a set of crucial molecular and cellular interactions in epithelial tissue cells, including effects due to DNA damage and mutation, to investigate the probability that cancer will \sout{occur} \tcb{progress} and, if so, when.

@@

\section{Methods}
We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo17_multiscale}. 
The agents in the model are the tissue cells \sout{with tumorigenic potential} \tcb{that have already formed a non-aggressive tumor yet have not progressed}.
\sout{These can either be mutation-free or, resulting from DNA damage during the cell cycle, can have any combination of three possible pathway mutations (shown in Fig. \ref{fig:ModelIntro}A).}
\tcb{
In the process of the simulation, these cells can acquire new pathway mutations among three possible (shown in Fig. \ref{fig:ModelIntro}A).
}
If the pathway is mutated, the component is colored red and is marked with `X'.
\sout{A single mutation is sufficient for the cell to be considered as a mutant cell.}
\tcb{
% I'm not a fan of "hypermutated", but I'll use it for now so that a quick replace all can change it later.
We distinguish between cells that have zero of the three pathway mutations and those that have at least one by referring to the latter as hypermutated.
}
\par
We model immune cells as continuous variables, \sout{appropriate since at the level of immune cells it is not necessary to keep track of single cell effects.}
\tcb{
effectively assuming that the TME is well-mixed with regards to the immune infiltrate.
}
The cytokine TGF-$\beta$ is \sout{\tcr{continuous} within the TME, i.e. the system is well-mixed.}
\tcb{
also assumed to be well-mixed in the TME so that all tissue cells have equal access to it.
}
Tissue cells can take on either epithelial or mesenchymal phenotypes in a plastic manner: these phenotypes depend on both the TME and cell-intrinsic factors.
\sout{While the score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.}
\tcb{While the EMT score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.}
% Above line replaced
In Fig. \ref{fig:ModelIntro}A, the EMT labeling is given by the \sout{color} \tcb{shade of blue} of the cell.
Even though each cell does have an EMT score on a spectrum of values, only the labeling influences its interactions in the model.

\subsection{Tissue Cells}\label{TissueCells}
The tissue cells at all times have associated with them two important quantities: which combination of mutations they carry and an EMT score.
The model has three pathways that have the potential to mutate over the course of a simulation.
There is a proliferation pathway that when mutated increases the probability of the cell proliferating.
There is an apoptosis pathway that when mutated decreases the probability of the cell undergoing apoptosis.
There is an immune evasion pathway that when mutated decreases the probability that an immune cell can effectively clear the mutated cell.
The EMT score will be exposited in Section \ref{EMT}.

\subsection{The Immune System}\label{ImmuneSystem}
The immune system is modeled as having three active immune cell types: NKs, CTLs, and Tregs.
% We could justify the immune system not clearing "base" mutated cells by saying the immune surveillance provides a baseline control over these cells accounted for in their apoptosis. Or suggesting that we are interested in immune activity above and beyond the baseline requirement to keep this tumor in check.
The NKs and CTLs act on the system by recognizing \sout{mutated} \tcb{hypermutated} cells and clearing them.
Upon successful clearance, they themselves are then deactivated and removed from the immune population.
Tregs act on the system by suppressing the effector functions of NKs and CTLs, making it less likely they can clear \sout{mutated} \tcb{hypermutated} cells.
In addition to this, Tregs release TGF-$\beta$ which further shapes the TME by pushing tissues cells more towards a mesenchymal phenotype.

@@

\subsection{EMT}\label{EMT}
Each tissue cell has an EMT score between 0 and 1.
When the score is above a fixed threshold value, the cell is labeled as mesenchymal.
Otherwise, it is below the threshold and labeled as epithelial.
For the purpose of \tcb{parameter values} of the model, a cell in an epithelial state is considered in the base state, and one in a mesenchymal state will have \sout{some} \tcb{a subset} of its parameters updated.
In particular, mesenchymal cells benefit from increased immune evasion but suffer from decreased proliferation.
Note, that because the immune system can only clear \sout{mutated} \tcb{hypermutated} cells, the increased immune evasion only benefits \sout{mutated} \tcb{hypermutated} mesenchymal cells.
\par
\tcb{
In modeling EMT this way, we are assuming that the same factor, TGF-$\beta$, drives EMT both at initiation and through progression of cancer.
In addition, as per our model assumptions (see Table \ref{table:model_assumptions}), we choose to ignore other factors that are known to induce EMT such as hypoxia as well as the typical spatial organization of epithelial and mesenchymal cells as this pertains to the invasive front.
}
\subsection{Model Simulation}

\subsubsection{Initial conditions} 
Every model run starts with \sout{a fixed amount} \tcb{an amount} of \tcb{tissue} cells, $N_0$ \tcb{, determined by the set of parameters used for the run.
A set of warmup cycles are run to ensure arrival to this steady state.
During these warmup cycles, no mutations are allowed to occur the only immune cells present are NKs.}
\sout{However, different parameter values will lead to different steady states for the population of tissue cells.}
\sout{In light of this, a warmup period of ~1000 cell cycles is used in which time no mutations are allowed to happen.}
\sout{Thus, the only immune cell present is NKs.}
After the warmup cycles are complete, a mutagenic event is simulated in which all the tissue cells experience a random increase in their probability to mutate.
From then on, a cell which proliferates also either mutates or increases its probability of mutating later on.

@@

\paragraph{Immune Clearance}
For both NK clearance and CTL clearance, the weights are built with the same factors but have different parameter values for NK and CTLs.
First of all, the cell needs to be \sout{mutated} \tcb{hypermutated} ($\delta_{\text{MUT}}=1$).
Second, there is a Hill factor that captures the probability of an immune cell finding and interacting with the given tissue cell with EC50 term $K_1$.
Third, NKs and CTLs have their own efficacy parameters, $E_{\text{NK}}$ and $E_{\text{CTL}}$, which can be understood as the rate of immune clearance given an immune cell has found the mutated cell.
Fourth, there is a decreasing Hill factor based on the number of Treg cells present with EC50 term $K_2$.
Finally, there are two factors that proportionally decrease the weight of immune clearance depending on if the cell has an immune evasion mutation ($\delta_{\text{IE}}=1$) or if it is mesenchymal ($\zeta=1$) with respective decreases $\Delta_{\text{IE}}$ and $\Delta_{\text{MIE}}$.
In total, the weight of NK clearance is given by

@@

For CTLs and Tregs, they rely on \sout{mutant} \tcb{hypermutated} cells being cleared before they can be activated.
Let $N_{\text{MUT}}^*(k)$ represent the number of \sout{mutant} \tcb{hypermutated} cells cleared by the immune system during cell cycle $k$.
In addition, Treg recruitment is upregulated by TGF-$\beta$, which will be incorporated via a Hill function with EC50 term $K_4$.
We choose the following differential equations to govern the CTL and Treg populations:

@@

\begin{table}[H]
\begin{center}
 \begin{tabular}{|| c | c||} 
 \hline
 {\bf Name} & {\bf Description}  \\ [0.5ex] 
 \hline
 $p$ & proliferation rate of tissue cells \\ 
 \hline
 $d_C$  & death rate of tissue cells \\
 \hline
$\Delta_\text{MIE}$ &  mesenchymal immune evasion \\
 \hline
 $\Delta_\text{MGA}$ & mesenchymal growth arrest    \\
 \hline
  $\Delta_A$ & \tcr{mutant cells decreased apoptosis}  \\ % should this and the others like it be more clear in stating this is not for all hypermutated cells, but only those with apoptosis pathway mutation?
  \hline
  $\Delta_\text{IE}$ & mutant cells increased immune evasion  \\
  \hline
  $\Delta_P$ & mutant cells increased proliferation  \\
  \hline
 $K_0$ & EC50 term for negative feedback of tissue cells on own proliferation\\
 \hline
 $K_1$ & EC50 term for probability of NK cell finding mutant cell\\
 \hline
  $K_2$ & EC50 term for Treg inhibition of cytotoxic functions  \\
  \hline
  $K_3$ & EC50 term for how much TGF-$\beta$ each cell has \\
  \hline
  $K_4$ & EC50 term for TGF-$\beta$ activation of Tregs \\
  \hline
 $E_\text{NK}$ & rate of NKs clearing mutants  \\
  \hline
  $E_\text{CTL}$ & rate of CTLs clearing mutants \\
  \hline
  $\sigma_\text{NK}$ & NK source rate \\ 
  \hline
  $\sigma_\text{CTL}$ & CTL source rate per cleared \sout{mutant} \tcb{hypermutated} cell \\ 
  \hline
  $\sigma_\text{Treg}$ & Treg source rate per cleared \sout{mutant} \tcb{hypermutated} cell \\ 
  \hline
  $d_\text{NK}$ & NK death rate \\ 
  \hline
  $d_\text{CTL}$ & CTL death rate \\ 
  \hline
  $d_\text{Treg}$ & Treg death rate \\ 
  \hline
  $k_\text{EMT}$ & EMT/MET rate  \\
  \hline
  $\sigma$ & standard deviation of noise in TGF-$\beta$ each cell receives  \\
  \hline
 $\tau_\text{max}$ & max amount of TGF-$\beta$ any cell can receive \\
  \hline 
 $\tau_\text{MUT}$ & rate of TGF-$\beta$ production by mutant cells\\
  \hline
 $\tau_\text{Treg}$ & rate of TGF-$\beta$ production by Treg\\
  \hline
\end{tabular}
  \caption{The model parameter names and descriptions. Note that many of these values are affected by the inflammation state of the system.}
\end{center}
\end{table}

\subsubsection{Mutation and tumorigenesis}
% 3rd Referee did not like this terminology for a tumorigenic event. Should we change it or better explain it?
At the end of each cell cycle, the proportion of tissue cells that are \sout{mutated} \tcb{hypermutated} is calculated, and if it is above a certain threshold, \sout{a tumorigenic event is recorded and the patient is determined to have cancer.}
\tcb{the simulation ends and we declare the tumor to have progressed.}
The Time to Cancer is given as the time from the start of the simulation, not including warmup, until \sout{tumorigenesis.} \tcb{this timepoint.}
Simulations run until either a \sout{tumorigenic event is recorded} \tcb{this occurs} or until the \tcb{predetermined} maximum number of cycles has been reached.

@@

\begin{figure}
\center
{\includegraphics[width=0.9\textwidth]{Figure1/Figure1.pdf}}
\caption{A. Schematic depiction of agent-based model components; \tcb{
each of the 10 columns represents a single tissue cell divided into three compartments representing the state (mutated or not) of the three pathways with mutagenic potential;
}\sout{blue/red denote tissue cells with/without mutations} \tcb{red/blue denotes mutated/unmutated pathways} . Black arrows depict regulation of the cell fate in each cell cycle. Inset depicts major interactions between the immune system and tissue cells.
 %The parameters used in this model can be found in the Supplementary material. Unless otherwise stated, parameter values are assumed to be those found in the Supplement. \tcr{Thoughts on the wording here? Once I have the table, I will use the name of that table.}
B. A representative simulation of one patient. The model parameter values used can be found in Supplementary Table 2. 
The inflammation cycling scheme is represented above the patient dynamics. The vertical dashed line denotes the end of the warmup period. Mut: \sout{mutated} \tcb{hypermutated} cells; Mes: mesenchymal cells. 
C. Survival curve for one cohort of patients with the parameter values given in Supplementary Table 2. }
\label{fig:ModelIntro}
\end{figure}

We begin by investigating general features of the model to establish baseline conditions and to assess the impact of various model components on the key measured outcomes: the \tcr{probability} of cancer, and the Time to Cancer. 

@@

The different means by which the immune system acts on tissue cells are also shown (\ref{fig:ModelIntro}A Inset). NK cells and CTLs attempt to clear \sout{mutated} \tcb{hypermutated} tissue cells, and deactivate upon successfully carrying out their cytotoxic function.

@@

After about 700 cell cycles, the hypermutated cell population begins to grow steadily.

@@

Considering the immune system dynamics, we see that the NK population is approximately constant, while the adaptive populations (CTL and Treg cells) grow quickly and dwarf NK cells in number following the accumulation of \sout{oncogenic mutations} \tcb{hypermutations}.

@@

When a cell transitions from an epithelial to a mesenchymal state, two phenotypic cell characteristics change: mesenchymal immune evasion (MIE) and mesenchymal growth arrest (MGA).
Both parameters are defined as proportions, and lie in $[0,1]$, where higher values indicate more mesenchymal-like properties.
The parameter $\Delta_\text{MIE}$ is the proportional reduction in the probability that a \sout{mutated} \tcb{hypermutated} mesenchymal cell will be subject to immune clearance in a given cycle.
The parameter $\Delta_\text{MGA}$ is the proportional reduction in the probability that a mesenchymal cell will proliferate in a given cell cycle, thus is equivalent (strictly in terms of its impact on the cell cycle) to an increased proportion of time spent resting in the $G_0$ phase.
The cytokine TGF-$\beta$ is also involved in the EMT process (by increasing the probability of EMT). All three of these parameters were found to have large effects on the model by the Morris OAT analysis presented in Section \ref{SensAnalysis}.
\par 
As MIE is increased, the Time to Cancer decreases (Fig. \ref{fig:FirstSurvivalCurves}A) under all sets of parameters studied: as this subpopulation of \sout{potentially mutated} \tcb{hypermutated} cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
These results are summarized by Fig. \ref{fig:FirstSurvivalCurves}A.
\par
Second, as MGA increases, Time to Cancer increases, i.e. lower proliferation rates for mesenchymal cells slow down cancer growth (Fig \ref{fig:FirstSurvivalCurves}B).
This is not immediately intuitive, since the decreased proliferation rate affects \sout{both mutated and non-mutated tissue cells.} \tcb{all mesenchymal cells, not just hypermutated cells.}
\par
Third, TGF-$\beta$ can be varied in two ways: the production by mesenchymal cells and the production by Treg cells.
In Fig \ref{fig:FirstSurvivalCurves}C, the results of varying Treg TGF-$\beta$ production are shown, indicating that an increased Treg TGF-$\beta$ production leads to a shorter Time to Cancer.
The two main ways in which TGF-$\beta$ influences the system is in recruitment of Treg cells and in pushing tissue cells to a mesenchymal phenotype.
Treg cells are modeled as tumor-protective and thus increasing their number will naturally decrease Time to Cancer.
Mesenchymal cells are more likely to evade the immune system, so pushing the system towards an overall more mesenchymal phenotype will better protect the cancer and decrease the Time to Cancer.

\begin{figure}
\center
{\includegraphics[width=0.95\textwidth]{Figure4/Figure4.pdf}}
\caption{Effects of inflammation on the time to cancer under different cycling schemes. A-B. As MIE varies, survival curves (each of 200 patients)  and corresponding bar plots to summarize the mean Time to Cancer for each cohort are shown. C-D. As MGA varies, survival curves  and corresponding bar plots to summarize the mean time to cancer for each cohort are shown.}
\label{fig:VaryINFL_and_MesPars}
\end{figure}

@@

Despite the intense interest in cancer and the immune system, and in the effects of EMT on cancer, there has not previously, to the best of our knowledge, been a model developed that combines these components: cancer, the immune system, and EMT. We saw this as an even more pressing need given the shared factors that influence all of these, such as TGF-$\beta$. We used an individual cell-based model framework to describe the multiscale processes that can lead to cancer: DNA damage occurs during the cell cycle and this can lead to mutations in pathways that affect the fitness of the cell, which in turn affects the cell population dynamics, which are also influenced by the intrinsic state of the cell (EMT), and extrinsic immune factors (inflammation, cytotoxic cells, etc).

@@

To test these predictions, we performed unsupervised analysis of pancreatic cancer data from The Cancer Genome Atlas, and looked at survival across two groups: we found that the presence of an EMT signature increased differences in survival between groups in combination with inflammation. We chose to summarize {\em in silico} patient simulations with a single parameter: the Time to Cancer, which we found captured the essential characteristics of the model. There are, of course, many tissue trajectories that all result in cancer, and it is our opinion that analysis of the transient cell dynamics in pre-cancerous tissues is a pressing need to shed insight into cellular biomarkers of cancer.

@@

These results represent promising steps in understanding the competing roles of the immune system and EMT during (pre-) development of epithelial cancers, yet much remains to be done. Further development of the inflammation module of this model is important given the large and sometimes paradoxical roles that the inflammatory state exerts on the epithelia and cancer-free survival (Figs. \ref{fig:MOAT} and \ref{fig:VaryINFL_and_MesPars}B, D). Currently, inflammation is modeled as independently cycling between high and low schemes, however many of the agents considered in the model actively contribute to the inflammatory state, thus it would be interesting to consider connecting these components (for example by letting the level of inflammation be conditional on the number of mutant cells in the tissue). Another layer of complexity is revealed by the natural anti-inflammatory role of Tregs. One consequence of the current model is that decreasing the number of Tregs increases the cancer-free survival. Clearly, there exists a trade-off to be accounted for, and adding to the model the main effector function of Tregs could remedy this and add depth to our understanding of the various roles that Tregs play in and around the tumor.

>>>>

In addition, the immune system plays important regulatory roles in \sout{pre-cancerous} \tcb{premalignant} tissue, where inflammation can increase the probability of tumorigenesis (and/or decrease the time to cancer), \sout{particularly} \tcb{including} in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.

@@

\tcb{There is evidence directly linking Treg-secreted TGF-$\beta$ with EMT in hepatocellular carcinoma\cite{shi2019cd4+}.}
Thus, even by considering only a single signaling factor, we find these three components (the tumor, the immune system, and EMT) to be linked. It therefore strikes us as a priority to develop models to understand how interactions between each of these three components affect cancer incidence and progression.

@@

We seek to describe a set of crucial molecular and cellular interactions in epithelial tissue cells, including effects due to DNA damage and mutation, to investigate the probability that \sout{cancer} \tcb{premalignant tissue} will \sout{occur} \tcb{progress} and, if so, when.

@@

\section{Methods}
We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo17_multiscale}. 
% might want to define "driver mutation" in intro. See PreMalignant_Thoughts.txt for definition found in a source
The agents in the model are the tissue cells \sout{with tumorigenic potential} \tcb{that have already formed a premalignant tumor yet lack key driver mutations}.
\sout{These can either be mutation-free or, resulting from DNA damage during the cell cycle, can have any combination of three possible pathway mutations (shown in Fig. \ref{fig:ModelIntro}A).}
\tcb{
In the process of the simulation, these cells can acquire driver mutations in any of three key pathways (shown in Fig. \ref{fig:ModelIntro}A).
}
If the pathway is mutated, the component is colored red and is marked with `X'.
\sout{A single mutation is sufficient for the cell to be considered as a mutant cell.}
\tcb{
% I'm not a fan of "malignant", but I'll use it for now so that a quick replace all can change it later.
We distinguish between cells that have zero driver mutations and those with at least one as premalignant and malignant, respectively.
}
\par
We model immune cells as continuous variables, \sout{appropriate since at the level of immune cells it is not necessary to keep track of single cell effects.}
\tcb{
effectively assuming that the TME is well-mixed with regards to the immune infiltrate.
}
The cytokine TGF-$\beta$ is \sout{\tcr{continuous} within the TME, i.e. the system is well-mixed.}
\tcb{
also assumed to be well-mixed in the TME so that all tissue cells have equal access to it.
}
Tissue cells can take on either epithelial or mesenchymal phenotypes in a plastic manner: these phenotypes depend on both the TME and cell-intrinsic factors.
\sout{While the score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.}
\tcb{While the EMT score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.}
% Above line replaced
In Fig. \ref{fig:ModelIntro}A, the EMT labeling is given by the \sout{color} \tcb{shade of blue} of the cell.
Even though each cell does have an EMT score on a spectrum of values, only the labeling influences its interactions in the model.

\subsection{Tissue Cells}\label{TissueCells}
The tissue cells at all times have associated with them two important quantities: which combination of \tcb{driver} mutations they carry and an EMT score.
The model has three pathways that have the potential to mutate over the course of a simulation.
There is a proliferation pathway that when mutated increases the probability of the cell proliferating.
There is an apoptosis pathway that when mutated decreases the probability of the cell undergoing apoptosis.
There is an immune evasion pathway that when mutated decreases the probability that an immune cell can effectively clear the mutated cell.
The EMT score will be exposited in Section \ref{EMT}.

\subsection{The Immune System}\label{ImmuneSystem}
The immune system is modeled as having three active immune cell types: NKs, CTLs, and Tregs.
% We could justify the immune system not clearing "base" mutated cells by saying the immune surveillance provides a baseline control over these cells accounted for in their apoptosis. Or suggesting that we are interested in immune activity above and beyond the baseline requirement to keep this tumor in check.
The NKs and CTLs act on the system by recognizing \sout{mutated} \tcb{malignant} cells and clearing them.
Upon successful clearance, they themselves are then deactivated and removed from the immune population.
Tregs act on the system by suppressing the effector functions of NKs and CTLs, making it less likely they can clear \sout{mutated} \tcb{malignant} cells.
In addition to this, Tregs release TGF-$\beta$ which further shapes the TME by pushing tissues cells more towards a mesenchymal phenotype.

@@

\subsection{EMT}\label{EMT}
Each tissue cell has an EMT score between 0 and 1.
When the score is above a fixed threshold value, the cell is labeled as mesenchymal.
Otherwise, it is below the threshold and labeled as epithelial.
For the purpose of \tcb{parameter values} of the model, a cell in an epithelial state is considered in the base state, and one in a mesenchymal state will have \sout{some} \tcb{a subset} of its parameters updated.
In particular, mesenchymal cells benefit from increased immune evasion but suffer from decreased proliferation.
Note, that because the immune system can only clear \sout{mutated} \tcb{malignant} cells, the increased immune evasion only benefits \sout{mutated} \tcb{malignant} mesenchymal cells.
\par
\tcb{
In modeling EMT this way, we are assuming that the same factor, TGF-$\beta$, drives EMT both at initiation and through progression of cancer.
In addition, as per our model assumptions (see Table \ref{table:model_assumptions}), we choose to ignore other factors that are known to induce EMT such as hypoxia as well as the typical spatial organization of epithelial and mesenchymal cells as this pertains to the invasive front.
}
\subsection{Model Simulation}

\subsubsection{Initial conditions} 
Every model run starts with \sout{a fixed amount} \tcb{an amount} of \tcb{premalignant tissue} cells, $N_0$ \tcb{, determined by the set of parameters used for the run.
A set of warmup cycles are run to ensure arrival to this steady state.
During these warmup cycles, no mutations are allowed to occur the only immune cells present are NKs.}
\sout{However, different parameter values will lead to different steady states for the population of tissue cells.}
\sout{In light of this, a warmup period of ~1000 cell cycles is used in which time no mutations are allowed to happen.}
\sout{Thus, the only immune cell present is NKs.}
After the warmup cycles are complete, a mutagenic event is simulated in which all the tissue cells experience a random increase in their probability to mutate.
From then on, a cell which proliferates also either mutates or increases its probability of mutating later on.

@@

\paragraph{Immune Clearance}
For both NK clearance and CTL clearance, the weights are built with the same factors but have different parameter values for NK and CTLs.
First of all, the cell needs to be \sout{mutated} \tcb{malignant} ($\delta_{\text{MUT}}=1$).
Second, there is a Hill factor that captures the probability of an immune cell finding and interacting with the given tissue cell with EC50 term $K_1$.
Third, NKs and CTLs have their own efficacy parameters, $E_{\text{NK}}$ and $E_{\text{CTL}}$, which can be understood as the rate of immune clearance given an immune cell has found the mutated cell.
Fourth, there is a decreasing Hill factor based on the number of Treg cells present with EC50 term $K_2$.
Finally, there are two factors that proportionally decrease the weight of immune clearance depending on if the cell has an immune evasion mutation ($\delta_{\text{IE}}=1$) or if it is mesenchymal ($\zeta=1$) with respective decreases $\Delta_{\text{IE}}$ and $\Delta_{\text{MIE}}$.
In total, the weight of NK clearance is given by

@@

For CTLs and Tregs, they rely on \sout{mutant} \tcb{malignant} cells being cleared before they can be activated.
Let $N_{\text{MUT}}^*(k)$ represent the number of \sout{mutant} \tcb{malignant} cells cleared by the immune system during cell cycle $k$.
In addition, Treg recruitment is upregulated by TGF-$\beta$, which will be incorporated via a Hill function with EC50 term $K_4$.
We choose the following differential equations to govern the CTL and Treg populations:

@@

\begin{table}[H]
\begin{center}
 \begin{tabular}{|| c | c||} 
 \hline
 {\bf Name} & {\bf Description}  \\ [0.5ex] 
 \hline
 $p$ & proliferation rate of tissue cells \\ 
 \hline
 $d_C$  & death rate of tissue cells \\
 \hline
$\Delta_\text{MIE}$ &  mesenchymal immune evasion \\
 \hline
 $\Delta_\text{MGA}$ & mesenchymal growth arrest    \\
 \hline
  $\Delta_A$ & \tcr{mutant cells decreased apoptosis}  \\ % should this and the others like it be more clear in stating this is not for all malignant cells, but only those with apoptosis pathway mutation?
  \hline
  $\Delta_\text{IE}$ & mutant cells increased immune evasion  \\
  \hline
  $\Delta_P$ & mutant cells increased proliferation  \\
  \hline
 $K_0$ & EC50 term for negative feedback of tissue cells on own proliferation\\
 \hline
 $K_1$ & EC50 term for probability of NK cell finding mutant cell\\
 \hline
  $K_2$ & EC50 term for Treg inhibition of cytotoxic functions  \\
  \hline
  $K_3$ & EC50 term for how much TGF-$\beta$ each cell has \\
  \hline
  $K_4$ & EC50 term for TGF-$\beta$ activation of Tregs \\
  \hline
 $E_\text{NK}$ & rate of NKs clearing mutants  \\
  \hline
  $E_\text{CTL}$ & rate of CTLs clearing mutants \\
  \hline
  $\sigma_\text{NK}$ & NK source rate \\ 
  \hline
  $\sigma_\text{CTL}$ & CTL source rate per cleared \sout{mutant} \tcb{malignant} cell \\ 
  \hline
  $\sigma_\text{Treg}$ & Treg source rate per cleared \sout{mutant} \tcb{malignant} cell \\ 
  \hline
  $d_\text{NK}$ & NK death rate \\ 
  \hline
  $d_\text{CTL}$ & CTL death rate \\ 
  \hline
  $d_\text{Treg}$ & Treg death rate \\ 
  \hline
  $k_\text{EMT}$ & EMT/MET rate  \\
  \hline
  $\sigma$ & standard deviation of noise in TGF-$\beta$ each cell receives  \\
  \hline
 $\tau_\text{max}$ & max amount of TGF-$\beta$ any cell can receive \\
  \hline 
 $\tau_\text{MUT}$ & rate of TGF-$\beta$ production by mutant cells\\
  \hline
 $\tau_\text{Treg}$ & rate of TGF-$\beta$ production by Treg\\
  \hline
\end{tabular}
  \caption{The model parameter names and descriptions. Note that many of these values are affected by the inflammation state of the system.}
\end{center}
\end{table}

\subsubsection{Mutation and tumorigenesis}
% 3rd Referee did not like this terminology for a tumorigenic event. Should we change it or better explain it?
At the end of each cell cycle, the proportion of tissue cells that are \sout{mutated} \tcb{malignant} is calculated, and if it is above a certain threshold, \sout{a tumorigenic event is recorded and the patient is determined to have cancer.}
\tcb{the simulation ends and we declare the tumor to have progressed.}
The Time to Cancer is given as the time from the start of the simulation, not including warmup, until \sout{tumorigenesis.} \tcb{this timepoint.}
Simulations run until either a \sout{tumorigenic event is recorded} \tcb{this occurs} or until the \tcb{predetermined} maximum number of cycles has been reached.

@@

\begin{figure}
\center
{\includegraphics[width=0.9\textwidth]{Figure1/Figure1.pdf}}
\caption{A. Schematic depiction of agent-based model components; \tcb{
each of the 10 columns represents a single tissue cell divided into three compartments representing the state (mutated or not) of the three pathways with mutagenic potential;
}\sout{blue/red denote tissue cells with/without mutations} \tcb{red/blue denotes mutated/unmutated pathways} . Black arrows depict regulation of the cell fate in each cell cycle. Inset depicts major interactions between the immune system and tissue cells.
 %The parameters used in this model can be found in the Supplementary material. Unless otherwise stated, parameter values are assumed to be those found in the Supplement. \tcr{Thoughts on the wording here? Once I have the table, I will use the name of that table.}
B. A representative simulation of one patient. The model parameter values used can be found in Supplementary Table 2. 
The inflammation cycling scheme is represented above the patient dynamics. The vertical dashed line denotes the end of the warmup period. Mut: \sout{mutated} \tcb{malignant} cells; Mes: mesenchymal cells. 
C. Survival curve for one cohort of patients with the parameter values given in Supplementary Table 2. }
\label{fig:ModelIntro}
\end{figure}

We begin by investigating general features of the model to establish baseline conditions and to assess the impact of various model components on the key measured outcomes: the \tcr{probability} of \sout{cancer} \tcb{progression}, and the Time to Cancer. 

@@

The different means by which the immune system acts on tissue cells are also shown (\ref{fig:ModelIntro}A Inset). NK cells and CTLs attempt to clear \sout{mutated} \tcb{malignant} tissue cells, and deactivate upon successfully carrying out their cytotoxic function.

@@

After about 700 cell cycles, the malignant cell population begins to grow steadily.

@@

Considering the immune system dynamics, we see that the NK population is approximately constant, while the adaptive populations (CTL and Treg cells) grow quickly and dwarf NK cells in number following the accumulation of \sout{oncogenic mutations} \tcb{driver mutations}.

@@

When a cell transitions from an epithelial to a mesenchymal state, two phenotypic cell characteristics change: mesenchymal immune evasion (MIE) and mesenchymal growth arrest (MGA).
Both parameters are defined as proportions, and lie in $[0,1]$, where higher values indicate more mesenchymal-like properties.
The parameter $\Delta_\text{MIE}$ is the proportional reduction in the probability that a \sout{mutated} \tcb{malignant} mesenchymal cell will be subject to immune clearance in a given cycle.
The parameter $\Delta_\text{MGA}$ is the proportional reduction in the probability that a mesenchymal cell will proliferate in a given cell cycle, thus is equivalent (strictly in terms of its impact on the cell cycle) to an increased proportion of time spent resting in the $G_0$ phase.
The cytokine TGF-$\beta$ is also involved in the EMT process (by increasing the probability of EMT). All three of these parameters were found to have large effects on the model by the Morris OAT analysis presented in Section \ref{SensAnalysis}.
\par 
As MIE is increased, the Time to Cancer decreases (Fig. \ref{fig:FirstSurvivalCurves}A) under all sets of parameters studied: as this subpopulation of \sout{potentially mutated} \tcb{malignant} cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
These results are summarized by Fig. \ref{fig:FirstSurvivalCurves}A.
\par
Second, as MGA increases, Time to Cancer increases, i.e. lower proliferation rates for mesenchymal cells slow down cancer growth (Fig \ref{fig:FirstSurvivalCurves}B).
This is not immediately intuitive, since the decreased proliferation rate affects \sout{both mutated and non-mutated tissue cells.} \tcb{all mesenchymal cells, not just malignant cells.}
\par
Third, TGF-$\beta$ can be varied in two ways: the production by mesenchymal cells and the production by Treg cells.
In Fig \ref{fig:FirstSurvivalCurves}C, the results of varying Treg TGF-$\beta$ production are shown, indicating that an increased Treg TGF-$\beta$ production leads to a shorter Time to Cancer.
The two main ways in which TGF-$\beta$ influences the system is in recruitment of Treg cells and in pushing tissue cells to a mesenchymal phenotype.
Treg cells are modeled as tumor-protective and thus increasing their number will naturally decrease Time to Cancer.
Mesenchymal cells are more likely to evade the immune system, so pushing the system towards an overall more mesenchymal phenotype will better protect the cancer and decrease the Time to Cancer.

\begin{figure}
\center
{\includegraphics[width=0.95\textwidth]{Figure4/Figure4.pdf}}
\caption{Effects of inflammation on the time to cancer under different cycling schemes. A-B. As MIE varies, survival curves (each of 200 patients)  and corresponding bar plots to summarize the mean Time to Cancer for each cohort are shown. C-D. As MGA varies, survival curves  and corresponding bar plots to summarize the mean Time to Cancer for each cohort are shown.}
\label{fig:VaryINFL_and_MesPars}
\end{figure}

@@

Despite the intense interest in cancer and the immune system, and in the effects of EMT on cancer, there has not previously, to the best of our knowledge, been a model developed that combines these \tcb{three} components\sout{: cancer, the immune system, and EMT.} \tcb{--cancer, the immune system, and EMT--during the progression from a premalignant to a malignant tumor.} We saw this as an even more pressing need given the shared factors that influence all of these, such as TGF-$\beta$. We used an individual cell-based model framework to describe the multiscale processes that can lead to cancer: DNA damage occurs during the cell cycle and this can lead to mutations in pathways that affect the fitness of the cell, which in turn affects the cell population dynamics, which are also influenced by the intrinsic state of the cell (EMT), and extrinsic immune factors (inflammation, cytotoxic cells, etc).

@@

To test these predictions, we performed unsupervised analysis of pancreatic cancer data from The Cancer Genome Atlas, and looked at survival across two groups: we found that the presence of an EMT signature increased differences in survival between groups in combination with inflammation. We chose to summarize {\em in silico} patient simulations with a single parameter: the Time to Cancer, which we found captured the essential characteristics of the model. There are, of course, many tissue trajectories that all result in cancer, and it is our opinion that analysis of the transient cell dynamics in \sout{pre-cancerous} \tcb{premalignant} tissues is a pressing need to shed insight into cellular biomarkers of cancer.

@@

These results represent promising steps in understanding the competing roles of the immune system and EMT during \sout{(pre-) development} \tcb{progression} of epithelial cancers, yet much remains to be done. Further development of the inflammation module of this model is important given the large and sometimes paradoxical roles that the inflammatory state exerts on the epithelia and cancer-free survival (Figs. \ref{fig:MOAT} and \ref{fig:VaryINFL_and_MesPars}B, D). Currently, inflammation is modeled as independently cycling between high and low schemes, however many of the agents considered in the model actively contribute to the inflammatory state, thus it would be interesting to consider connecting these components (for example by letting the level of inflammation be conditional on the number of mutant cells in the tissue). Another layer of complexity is revealed by the natural anti-inflammatory role of Tregs. One consequence of the current model is that decreasing the number of Tregs increases the cancer-free survival. Clearly, there exists a trade-off to be accounted for, and adding to the model the main effector function of Tregs could remedy this and add depth to our understanding of the various roles that Tregs play in and around the tumor.
